---
title: "EDA Technique Summary"
author: "Group 3"
date: "May 15, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Technique 1: The Variation of Continious and Categorical Variables (Alireza Mostafizi)
This chapter briefly explaines a few tehcniques to visualized the distribution of the categorical and continious variables. *Categorical Values* are normally visualized with bar plots (*geom_bar*) and *continious values* are typically visualized with histograms (*geom_histogram* or *geom_freqpoly*). The following code plots the distribution of *cut* (categorical) and *carat* (continious) variables from *dimonds* dataset in *ggplot2*.

```{r, message=FALSE}
library(ggplot2, tidyverse)
library(gridExtra)
library(dplyr)
# Categorical Variable
cut_dist <- ggplot(data = diamonds) +
            geom_bar(mapping = aes(x = cut))
# Continious Variable
carat_dist <- ggplot(data = diamonds) +
              geom_histogram(mapping = aes(x = carat), binwidth = 0.5)
grid.arrange(cut_dist, carat_dist, ncol = 2)
```

Please note the use of *grid.arrange()* function from the package *ggExtra*. Also, I recommend always trying different *bin_width*. Small *bin_disth* helps you find the most common values in a better way as folllwing,

```{r}
ggplot(data = diamonds, mapping = aes(x = carat)) +
  geom_histogram(binwidth = 0.01) # Low bin_width reveals the common values that
                                  # were cluttered in the original histogram
```

There might be some small details hidden in large *bin_widths*. In addition, if, for some reason, you rather using any other type of plot, you can generate the dataset with the count values with the following code for both categorical and continious variables.

```{r}
library(dplyr) ## Needed for the pip function
# categorical variable
diamonds %>% ## %>% is the pipe function. Similar to | in unix systems
  count(cut)
# continous variable
diamonds %>% 
  count(cut_width(carat, 0.5)) # binwidth of 0.5
```
Anothr usefull technique for overlaying different historgrams is to use *geom_freqpoly()* instead of *geom_histogram()*. It is exactly the same but instead of bars, a poly line represents the histogram.


```{r}
ggplot(data = diamonds, mapping = aes(x = carat, colour = cut)) +
  geom_freqpoly(binwidth = 0.1)
```

Another good technique is zooming into a plot to find the unusual values in a distribution. *coord_cartesian()* function and its parameters, *xlim* and *ylim* come in handy for this purpose. Let's try it on the histogram of *y*, the width of the dimonds.

```{r}
# Normal histogram
y_dist <- ggplot(data = diamonds) +
          geom_histogram(mapping = aes(x = y), binwidth = 0.5)
# Zoomed in histogran
y_dist_zoomed <- ggplot(data = diamonds) +
                 geom_histogram(mapping = aes(x = y), binwidth = 0.5) +
                 coord_cartesian(ylim = c(0,5))
grid.arrange(y_dist, y_dist_zoomed, ncol = 2)
```
That clears the unusual values in 0, around 30 and around 60. Alternatively, you can found these observations with *filter()* function from *dplyr* package.

```{r}
diamonds %>% 
  filter(y < 3 | y > 20) %>% 
  select(price, x, y, z) %>%
  arrange(y)
```
