---
title: "Alternative Commute Plot"
author: "Group 3"
date: "April 19, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Figure 1 shows the alternative commute plot that is made specifically to emphasize on the comparisons of interest, i.e., to show that people in Oregon commute by bicycle more than the other neighboring states. The improvements over the original plot are as following.

1. The first improvement is that the plots are narrowed down to specifically represent the bicycle mode share, not all other modes as well.

2. The highest perceptual task, **position on a common scale**, has been employed as opposed to the angle and area in the original plot.

3. The comparisons are kept close to eachother in one plot, as opposed to the original plot in which the observer had to take 6 different values to compare from 6 different plots.

4. To emphasize on the comparison between Oregon and all other states, the associated value is colored differently than all others.

```{r alt_plot, echo=FALSE, message=FALSE, fig.cap="\\label{fig:alt_plot} Alternative Commute Plot"}
library(tidyverse, ggplot2)

commute_nw_bike <- structure(list(JWTR = c(9L, 9L, 9L, 9L, 9L, 9L), transport_type = c("Bicycle", 
"Bicycle", "Bicycle", "Bicycle", "Bicycle", "Bicycle"), avg_time = c(16.5704697986577, 
20.0443760420675, 15.801775147929, 22.6850393700787, 21.4092872570194, 
24.1366459627329), med_time = c(15, 15, 12, 16.5, 20, 20), min_time = c(1L, 
1L, 1L, 1L, 1L, 1L), max_time = c(60L, 178L, 80L, 155L, 183L, 
154L), q25_time = c(5, 10, 5.25, 10, 10, 10), q75_time = c(20, 
25, 20, 30, 30, 30), n = c(149L, 7797L, 338L, 254L, 1852L, 1288L
), n_missing = c(0L, 0L, 0L, 0L, 0L, 0L), state = c("ak", "ca", 
"id", "nv", "or", "wa"), state_n = c(14570L, 769646L, 33210L, 
58313L, 83168L, 152527L), prop = c(0.0102264927934111, 0.0101306314851243, 
0.0101776573321289, 0.00435580402311663, 0.0222681800692574, 
0.00844440656408374), state_name = c("Alaska", "California", 
"Idaho", "Nevada", "Oregon", "Washington")), class = c("grouped_df", 
"tbl_df", "tbl", "data.frame"), row.names = c(NA, -6L), vars = "state", .Names = c("JWTR", 
"transport_type", "avg_time", "med_time", "min_time", "max_time", 
"q25_time", "q75_time", "n", "n_missing", "state", "state_n", 
"prop", "state_name"), indices = list(0L, 1L, 2L, 3L, 4L, 5L), group_sizes = c(1L, 
1L, 1L, 1L, 1L, 1L), biggest_group_size = 1L, labels = structure(list(
state = c("ak", "ca", "id", "nv", "or", "wa")), class = "data.frame", row.names = c(NA, 
-6L), vars = "state", .Names = "state"))


ggplot(data = commute_nw_bike) +
  geom_bar(mapping = aes(x = reorder(state_name, -prop), y = prop, fill = (state_name == "Oregon")), stat = "identity") +
  xlab("State") +
  ylab("Proportion") +
  guides(fill = FALSE) +
  theme_minimal()
```

